\documentclass{article}
\usepackage{proof}
\newcommand{\tal}{{\sf TALx86}}
\title{\tal\ Language Specification}
\setlength{\oddsidemargin}{0.25in}
\setlength{\evensidemargin}{0.25in}
\setlength{\textwidth}{6 in}
\author{Neal Glew}
\newenvironment{gramrule}
  {\begin{flushleft}$\begin{array}{p{0.7in}ll}}
  {\end{array}$\end{flushleft}}
\newcommand{\nterm}[1]{{\it#1}}
\newcommand{\term}[1]{{\it#1}}
\newcommand{\ts}[1]{\hbox{\tt#1}}
\newcommand{\alt}{\mathrel{|}}
\begin{document}
\maketitle
There are two kinds of \tal\ compilation units: interfaces and
implementations.  The \tal\ compiler converts finite sequences of ASCII
characters into \tal\ compilation units as follows.  First it tokenises the
character sequence, then it parses the token sequence, and finally it performs
some context sensitive checks.  To tokenise the characters, the \tal\ compiler
starts with the first character of the sequence and finds the longest matching
token as defined in the lexical structure section.  Then it finds the longest
matching token starting at the character following the first token.  This
process continues until all characters have been tokenised or a character is
found that does not start a token.  The latter case is a lexical error.
Otherwise, whitespace and comment tokens are discarded and an $\term{eof}$
token is added to the end.  These tokens are the terminal symbols of the
grammar in the section on grammatical structure and the \tal\ compiler attempts
to find a parse tree with either $\nterm{int}$ or $\nterm{imp}$ as starting
symbol respectively for an interface or an implementation.
\section{Lexical Structure}
Literal ASCII characters are written in \ts{this} font, ASCII
character codes are written like this, $10$, and token names are
written in \term{this} font.  The characters \verb+~+, \verb+\+,
\ts{/}, \ts{\&}, \ts{\%}, \ts{\#}, and \ts{!}  are currently unused
but might be used in the future.  The characters 0--8, 14--31, and
127--255 are illegal.  The characters \ts{<}, \ts{>}, \ts{(}, \ts{)},
\ts{[}, \ts{]}, \verb+{+, \verb+}+, \ts{|}, \ts{@}, \ts{,}, \ts{.},
\ts{*}, \ts{+}, \ts{:}, \ts{=}, \verb+^+, and \ts{`} are tokens and the
character itself denotes the terminal symbol.  The character sequences
\ts{-!>}, \ts{::}, \ts{<=}, \verb+^r+, \verb+^w+, \verb+^rw+, and
\verb+^u+ are tokens and the sequences themselves denote the terminal
symbol.
\begin{gramrule}
\term{whitespace} &::=& (32 \alt 9 \alt 11 \alt 12)^+ \\
\term{eol} &::=& 10 \alt 13 \alt 10\;13 \alt 13\;10 \\
\term{comment} &::=& \ts{;} [\verb+~+\;10\;13]^* \; \term{eol} \\
\term{num} &::=& \ts{-}^? [\ts{0}-\ts{9}]^+(\ts{d} \alt \ts{t})^? \\
&|& [\ts{0}-\ts{9}]^+\;(\ts{b} \alt \ts{y} \alt \ts{o} \alt \ts{q}) \\
&|& [\ts{0}-\ts{9}][\ts{0}-\ts{9}\;\ts{a}-\ts{f}\;\ts{A}-\ts{F}]^*\;\ts{h} \\
\term{string} &::=& \ts{"} \; [32\;33\;35-\verb+~+]^* \; \ts{"}
\end{gramrule}
Identifiers are tokens that match the following rule and are not reserved
words.
\begin{gramrule}
\term{ident} &::=& [\ts{a}-\ts{z}\;\ts{A}-\ts{Z}\;\ts{\_}\;\ts{\$}\;\ts{?}]
[\ts{a}-\ts{z}\;\ts{A}-\ts{Z}\;\ts{\_}\;\ts{\$}\;\ts{?}\;\ts{0}-\ts{9}]^*
\end{gramrule}
There are three kinds of reserved words: case insensitive, case sensitive, and
filename keywords.

The case insensitive keywords are any capitilisation of the following
words, the lowercase version itself denotes the terminal symbol:
\ts{adc}, \ts{add}, \ts{al}, \ts{and}, \ts{asub}, \ts{aupd},
\ts{bl}, \ts{bswap}, \ts{btagi}, \ts{btagvar}, \ts{call},
\ts{cbw}, \ts{cdq}, \ts{cl}, \ts{clc}, \ts{cmc}, \ts{cmova},
\ts{cmovae}, \ts{cmovb}, \ts{cmovbe}, \ts{cmovc}, \ts{cmove},
\ts{cmovg}, \ts{cmovge}, \ts{cmovl}, \ts{cmovle}, \ts{cmovna},
\ts{cmovnae}, \ts{cmovnb}, \ts{cmovnbe}, \ts{cmovnc}, \ts{cmovne},
\ts{cmovng}, \ts{cmovnge}, \ts{cmovnl}, \ts{cmovnle}, \ts{cmovno},
\ts{cmovnp}, \ts{cmovns}, \ts{cmovnz}, \ts{cmovo}, \ts{cmovp},
\ts{cmovpe}, \ts{cmovpo}, \ts{cmovs}, \ts{cmovz}, \ts{cmp}, \ts{code},
\ts{coerce}, \ts{cwd}, \ts{cwde}, \ts{data}, \ts{db}, \ts{dd},
\ts{dec}, \ts{div}, \ts{dl}, \ts{dw}, \ts{dword}, \ts{eax}, \ts{ebp},
\ts{ebx}, \ts{ecx}, \ts{edi}, \ts{edx}, \ts{end}, \ts{esi}, \ts{esp},
\ts{fallthru}, \ts{idiv}, \ts{imul}, \ts{inc}, \ts{int}, \ts{into},
\ts{ja}, \ts{jae}, \ts{jb}, \ts{jbe}, \ts{jc}, \ts{je}, \ts{jecxz},
\ts{jg}, \ts{jge}, \ts{jl}, \ts{jle}, \ts{jmp}, \ts{jna}, \ts{jnae},
\ts{jnb}, \ts{jnbe}, \ts{jnc}, \ts{jne}, \ts{jng}, \ts{jnge},
\ts{jnl}, \ts{jnle}, \ts{jno}, \ts{jnp}, \ts{jns}, \ts{jnz}, \ts{jo},
\ts{jp}, \ts{jpe}, \ts{jpo}, \ts{js}, \ts{jz}, \ts{labeltype},
\ts{lahf}, \ts{lea}, \ts{loopd}, \ts{looped}, \ts{loopned},
\ts{malloc}, \ts{mov}, \ts{movsx}, \ts{movzx}, \ts{mul}, \ts{neg},
\ts{nop}, \ts{not}, \ts{or}, \ts{pop}, \ts{popad}, \ts{popfd},
\ts{ptr}, \ts{push}, \ts{pushad}, \ts{pushfd}, \ts{rcl}, \ts{rcr},
\ts{retn}, \ts{rol}, \ts{ror}, \ts{sahf}, \ts{sal}, \ts{sar},
\ts{sbb}, \ts{seta}, \ts{setae}, \ts{setb}, \ts{setbe}, \ts{setc},
\ts{sete}, \ts{setg}, \ts{setge}, \ts{setl}, \ts{setle}, \ts{setna},
\ts{setnae}, \ts{setnb}, \ts{setnbe}, \ts{setnc}, \ts{setne},
\ts{setng}, \ts{setnge}, \ts{setnl}, \ts{setnle}, \ts{setno},
\ts{setnp}, \ts{setns}, \ts{setnz}, \ts{seto}, \ts{setp}, \ts{setpe},
\ts{setpo}, \ts{sets}, \ts{setz}, \ts{shl}, \ts{shld}, \ts{shr},
\ts{shrd}, \ts{stc}, \ts{sub}, \ts{tal\_struct}, \ts{tal\_ends},
\ts{test}, \ts{type}, \ts{unpack}, \ts{val}, \ts{xchg}, and \ts{xor}.

The case sensitive keywords are the following character sequences, the sequence
itself denotes the terminal symbol: \ts{?}, \ts{\_begin\_TAL}, \ts{\_end\_TAL},
\ts{All}, \ts{array}, \ts{B1}, \ts{B2}, \ts{B4}, \ts{B8},
\ts{Exist}, \ts{fn}, \ts{junk},
\ts{pack}, \ts{R}, \ts{rec}, \ts{RL}, \ts{roll}, \ts{S},
\ts{se}, \ts{Sint}, \ts{slot}, \ts{sptr}, \ts{subsume}, \ts{sum}, \ts{T},
\ts{T1}, \ts{T2}, \ts{T4}, \ts{T8}, \ts{Tm}, \ts{Ts}, and \ts{unroll}.

There are three filename keywords given below:
\begin{gramrule}
\term{include} &::=& \ts{include}\;filename \\
\term{tal\_export} &::=& \ts{tal\_export}\;filename \\
\term{tal\_import} &::=& \ts{tal\_import}\;filename \\
\term{filename} &::=& [32\;9\;11\;12]^+ \; [33-126]^+ \\
\end{gramrule}

The grammar also refers to terminal symbol \term{cc} which is any
capatalisation of the character sequences: \ts{a}, \ts{ae}, \ts{b}, \ts{be},
\ts{c}, \ts{e}, \ts{g}, \ts{ge}, \ts{l}, \ts{le}, \ts{na}, \ts{nae}, \ts{nb},
\ts{nbe}, \ts{nc}, \ts{ne}, \ts{ng}, \ts{nge}, \ts{nl}, \ts{nle}, \ts{no},
\ts{np}, \ts{ns}, \ts{nz}, \ts{o}, \ts{p}, \ts{pe}, \ts{po}, \ts{s}, and
\ts{z}.  In places other than where \term{cc} appears in the grammar these
character sequences are just identifiers.
\newpage
\section{Grammatical Structure}
\tal's grammar follows.  $\alpha^\beta_\gamma$ means a list of $\beta$
$\alpha$s seperated by $\gamma$s; $\cdot^*$ means zero or more; $\cdot^+$ means
one or more; $\cdot^?$ means zero or one; $\alt$ means or.  The
number superscripts on alternatives indicate precedence 1 being lowest
100 being highest.  The superscripts $r$ and $l$ indicate right and
left associativity respectively.
\subsection{Compilation Units}
\begin{gramrule}
\nterm{int} &::=& (\nterm{intitem} \alt \term{eol})^* \; \term{eof} \\
\nterm{intitem} &::=& \term{include} \; \term{eol} \\
&|& \ts{type} \; \ts{<} \nterm{label} \ts{:} \nterm{kind} \ts{>}\;\term{eol} \\
&|& \ts{type} \; \ts{<} \nterm{label} \ts{:} \nterm{kind} \ts{=} \nterm{con}
     \ts{>}\;\term{eol} \\
&|& \ts{type} \; \ts{<} \nterm{label} \ts{:} \nterm{kind} \ts{<=} \nterm{con}
     \ts{>}\;\term{eol} \\
&|& \ts{val} \; \nterm{label} \ts{,} \ts{<} \nterm{con} \ts{>} \; \term{eol}
\end{gramrule}
Strictly speaking, the grammar is more liberal with than the rule given here:
Export declaration placement is less restricted.  Also the first
$\term{include}$ must be for the file ``{\tt tal.inc}''.
\begin{gramrule}
\nterm{imp} &::=& \term{eol}^* \; \term{include} \; \term{eol}^+ \\
&& \ts{\_begin\_TAL} \; \term{eol}^+ \\
&& (\term{tal\_import} \; \term{eol}^+)^* \\
&& (\term{tal\_export} \; \term{eol}^+)^* \\
&& (\ts{type} \; \ts< \nterm{label} \ts{=} \nterm{con} \ts{>} \;
   \term{eol}^+)^* \\ 
&& ( \; \ts{type} \; \ts< \nterm{label} \ts{:} \nterm{kind} \ts{=} \nterm{con}
     \ts{>} \; \term{eol}^+ \\
&& | \; \ts{code} \; \term{eol}^+ \\
&& \quad (\nterm{label} \ts{:} \; \term{eol}^* \\
&& \quad \phantom{(} \ts{labeltype} \ts{<} \nterm{con} \ts{>} \;\term{eol}^+ \\
&& \quad \phantom{(} (\nterm{instruction}\;\term{eol}^+)^* \\
&& \quad )^* \\
&& | \; \ts{data} \; \term{eol}^+ \\
&& \quad (\nterm{label} \ts{:} \; \term{eol}^* \\
&& \quad \phantom{(} (\ts{labeltype} \ts{<} \nterm{con} \ts{>} \;
                      \term{eol}^+)^? \\
&& \quad \phantom{(} (\ts{coerce} \ts{<} \nterm{coerce}[\ts{?}] \ts{>} \;
                      \term{eol}^+)^? \\
&& \quad \phantom{(} (\nterm{dataitem}\;\term{eol}^+)^* \\
&& \quad )^* \\
&& )^* \\
&& \ts{\_end\_TAL} \; \term{eol}^+ \\
&& \ts{end} \; \term{eol}^* \; \term{eof}
\end{gramrule}
\subsection{Instructions}
\begin{gramrule}
\nterm{instruction} &::=&
    (\ts{adc} \alt \ts{add} \alt \ts{and} \alt \ts{imul} \alt \ts{or} \alt
    \ts{sbb} \alt \ts{sub} \alt \ts{xor}) \; \nterm{binop} \\
&|& \ts{bswap} \; \nterm{reg} \\
&|& \ts{call} \; \nterm{unaryop\_c} \\
&|& \ts{cbw} \alt \ts{cdq} \alt \ts{cwd} \alt \ts{cwde} \\
&|& \ts{clc} \alt \ts{cmc} \alt \ts{stc} \\
&|& \ts{cmovcc} \; \nterm{reg} \ts{,} \nterm{unaryop\_c} \\
&|& (\ts{cmp} \alt \ts{test}) \; \nterm{binop} \\
&|& (\ts{dec} \alt \ts{div} \alt \ts{idiv} \alt \ts{imul} \alt \ts{inc} \alt
    \ts{mul} \alt \ts{neg} \alt \ts{not}) \; \nterm{unaryop} \\
&|& \ts{imul} \; \nterm{binaryop} \ts{,} \term{num} \\
&|& \ts{int} \; \term{num} \\
&|& \ts{into} \\
&|& (\ts{jcc} \alt \ts{jecxz} \alt \ts{loopd} \alt \ts{looped} \alt
    \ts{loopned}) \; \nterm{coerce}[\nterm{label}] \\
&|& \ts{jmp} \; \nterm{unaryop\_c} \\
&|& \ts{lea} \; \nterm{reg} \ts{,} \nterm{unaryop} \\
&|& \ts{lahf} \alt \ts{sahf} \\
&|& \ts{mov} \; \nterm{unaryop} \ts{,} \nterm{unaryop\_c} \\
&|& \ts{nop} \\
&|& \ts{pop} \; \nterm{unaryop} \\
&|& \ts{popad} \alt \ts{popfd} \\
&|& \ts{push} \; \nterm{unaryop\_c} \\
&|& \ts{pushad} \alt \ts{pushdfd} \\
&|& (\ts{rcl} \alt \ts{rcr} \alt \ts{rol} \alt \ts{ror} \alt \ts{sal} \alt
    \ts{sar} \alt \ts{shl} \alt \ts{shr}) \; \nterm{unaryop} \ts{,}
    \nterm{shift} \\
&|& \ts{retn} \; \term{num}^? \\
&|& \ts{setcc} \; \nterm{reglow} \\
&|& (\ts{shld} \alt \ts{shrd}) \; \nterm{unaryop} \ts{,} \nterm{reg} \ts{,}
    \nterm{shift} \\
&|& \ts{xchg} \; \nterm{unaryop} \ts{,} \nterm{reg} \\
\multicolumn{3}{l}{\hbox{TAL specific instructions}} \\
&|& \ts{asub} \; \nterm{reg} \ts{,}
    \nterm{coerce}[\nterm{reg}\alt\nterm{label}] (\ts{+} \term{num})^? \ts{,}
    \term{num} \ts{,} \nterm{reg} \ts{,} \nterm{genop} \\
&|& \ts{aupd} \; \nterm{coerce}[\nterm{reg}\alt\nterm{label}]
    (\ts{+} \term{num})^? \ts{,}
    \term{num} \ts{,} \nterm{reg} \ts{,} \nterm{reg} \ts{,} \nterm{genop} \\
&|& \ts{btagi} \; \term{cc} \ts{,} \nterm{reg} \ts{,} \term{num} \ts{,}
    \nterm{coerce}[\nterm{label}] \\
&|& \ts{btagvar} \; \term{cc} \ts{,} \ts{[} \nterm{reg} \ts{+}
    \term{num} \ts{]} \ts{,} \term{num} \ts{,} \nterm{coerce}[\nterm{label}] \\
&|& \ts{coerce} \; \nterm{coerce}[\nterm{reg}] \\
&|& \ts{fallthru} \; (\ts{<} \nterm{con}^*_{\ts{,}} \ts{>})^? \\
&|& \ts{malloc} \; \term{num} \ts{,} \nterm{mallocarg} (\ts{,}\ts{<}
    \nterm{con} \ts{>})^? \\
&|& \ts{unpack} \; \nterm{convar} \ts{,} \nterm{reg} \ts{,}
    \nterm{unaryop\_c} 
\end{gramrule}
\begin{gramrule}
\nterm{mallocarg} &::=& \ts{<} \nterm{mallocarg'} \ts{>} \\
\nterm{mallocarg'} &::=& \ts{[} \nterm{mallocarg'}_{\ts{,}}^* \ts{]} \\
&|& \ts{:} \nterm{con} \\
&|& \ts{array} \ts{(} \term{num} \ts{,} (\ts{B1} \alt \ts{B2} \alt
    \ts{B4} \alt \ts{B8}) \ts {)}
\end{gramrule}
\subsection{Operands}
\begin{gramrule}
\nterm{unaryop} &::=& (\ts{dword} \; \ts{ptr})^? \; \nterm{genop} \\
\nterm{unaryop\_c} &::=& (\ts{dword} \; \ts{ptr})^? \;
    \nterm{coerce}[\nterm{genop}] \\
\nterm{binop} &::=& \nterm{unaryop} \ts{,} \nterm{unaryop} \\
\end{gramrule}
\begin{gramrule}
\nterm{genop} &::=& \term{num} \\
&|& \ts{S} \ts{(} \term{num} \ts{)} \\
&|& \nterm{reg} \\
&|& \nterm{label} \\
&|& \ts{[} \nterm{coerce}[\nterm{reg}\alt\nterm{label}] (\ts{+} \term{num})^?
    \ts{]}
\end{gramrule}
\begin{gramrule}
\nterm{coerce}\hbox{$[\alpha]$} &::=& \alpha \\
&|& \ts{pack} \ts{(} \ts{<} \nterm{con} \ts{>} \ts{,} \nterm{coerce}[\alpha]
    \ts{,} \ts{<} \nterm{con} \ts{>} \ts{)} \\
&|& \ts{tapp} \ts{(} \nterm{coerce}[\alpha] \ts{,} \ts{<}
    \nterm{con}^+_{\ts{,}} \ts{>} \ts{)} \\
&|& \ts{roll} \ts{(} \ts{<} \nterm{con} \ts{>} \ts{,} \nterm{coerce}[\alpha]
    \ts{)} \\
&|& \ts{unroll} \ts{(} \nterm{coerce}[\alpha] \ts{)} \\
&|& \ts{sum} \ts{(} \ts{<} \nterm{con} \ts{>} \ts{,} \nterm{coerce}[\alpha]
    \ts{)} \\
&|& \ts{rollsum} \ts{(} \ts{<} \nterm{con} \ts{>} \ts{,} \nterm{coerce}[\alpha]
    \ts{)} \\
&|& \ts{rec} \ts{(} \nterm{coerce}[\alpha] \ts{)} \\
&|& \ts{array} \ts{(} \term{num} \ts{,} \term{num} \ts{,}
    \nterm{coerce}[\alpha] \ts{)} \\
&|& \ts{slot} \ts{(} \term{num} \ts{,} \term{num} \ts{,}
    \nterm{coerce}[\alpha] \ts{)} \\
&|& \ts{subsume} \ts{(} \ts{<} \nterm{con} \ts{>} \ts{,} \nterm{coerce}[\alpha]
    \ts{)} \\
\end{gramrule}
\begin{gramrule}
\nterm{reg} &::=& \ts{eax} \alt \ts{ebx} \alt \ts{ecx} \alt \ts{edx} \alt
   \ts{esi} \alt \ts{edi} \alt \ts{ebp} \alt \ts{esp} \alt \ts{R} \ts{(}
   \term{ident} \ts{)} \\
\nterm{reglow} &::=& \ts{al} \alt \ts{bl} \alt \ts{cl} \alt \ts{dl} \alt 
   \ts{RL} \ts{(} \term{ident} \ts{)} \\
\nterm{shift} &::=& \ts{cl} \alt \term{num}
\end{gramrule}
\subsection{Data}
\begin{gramrule}
\nterm{dataitem} &::=& \ts{db} \; (\term{num}\alt\term{string})^+_{\ts{,}} \\
&|& \ts{dw} \; \term{num} \\
&|& \ts{dd} \; \ts{?} \\
&|& \ts{dd} \; \ts{S} \ts{(} \term{num} \ts{)} \\
&|& \ts{dd} \; \term{num} \\
&|& \ts{tal\_struct} \\
&|& \ts{tal\_ends}
\end{gramrule}
\subsection{Type Constructors and Kinds}
\begin{gramrule}
\nterm{con} &::=& \ts{(} \nterm{con} \ts{)} \\
&|& \nterm{convar} \\
&|^1& \ts{fn} (\nterm{convar} \ts{:} \nterm{kind})^+ \ts{.} \nterm{con} \\
&|^{2l}& \nterm{con} \; \nterm{con} \\
&|& \ts{[} \nterm{con}^*_{\ts{,}} \ts{]} \\
&|^3& \nterm{con} \ts{.} \term{num} \\
&|& \ts{`} \nterm{label} \\
&|& \nterm{primcon} \\
&|& \ts{rec} \ts{(} (\nterm{convar} \ts{:} \nterm{kind} \ts{.}
    \nterm{con})^+_{\ts{,}} \ts{)} \\
&|^4& \ts{All} \ts{[} (\nterm{convar} \ts{:} \nterm{kind})^+ \ts{]} \ts{.} 
      \nterm{con} \\
&|^4& \ts{Exist} \ts{[} (\nterm{convar} \ts{:} \nterm{kind})^+ \ts{]} \ts{.} 
      \nterm{con} \\
&|& \verb+{+ (\nterm{reg} \ts{:} \nterm{con})^*_{\ts{,}} \verb+}+ \\
&|^4& \verb+^+ \ts{T} \ts{[} \nterm{num}^+_{\ts{,}} \ts{]} \\
&|^4& \verb+^+ ( \ts{T} \ts{(} \nterm{num}^+_{\ts{,}} \ts{)} )^? \nterm{con} \\
&|^4& \verb+^+ \ts{T} (\verb+^r+ \alt \verb+^w+ \alt \verb+^rw+ \alt
      \verb+^u+) \ts{(} \nterm{con} (\ts{,} \nterm{num})^* \ts{)}
      \nterm{con} \\
&|^7& \nterm{con} (\verb+^r+ \alt \verb+^w+ \alt \verb+^rw+ \alt \verb+^u+) \\
&|& \ts{*} \ts{[} \nterm{con}^*_{\ts{,}} \ts{]} \\
&|& \ts{+} \ts{[} \nterm{con}^*_{\ts{,}} \ts{]} \\
&|& \ts{array} \ts{(} \nterm{con} \ts{,} \nterm{con} \ts{)} \\
&|& \ts{S} \ts{(} \nterm{con} \ts{)} \\
&|^2& \ts{sptr} \; \nterm{con} \\
&|& \ts{se} \\
&|^{6r}& \nterm{con} \ts{::} \nterm{con} \\
&|^{5r}& \nterm{con} \ts{@} \nterm{con}
\end{gramrule}
\begin{gramrule}
\nterm{primcon} &::=& \ts{B1} \alt \ts{B2} \alt \ts{B4} \alt \ts{B8} \\
&|& \ts{junk} \; \term{num} \\
&|& \term{num}
\end{gramrule}
\begin{gramrule}
\nterm{kind} &::=& \ts{(} \nterm{kind} \ts{)} \\
&|& \ts{T} \\
&|& \ts{T1} \alt \ts{T2} \alt \ts{T4} \alt \ts{T8} \\
&|& \ts{Tm} \\
&|& \ts{Tm} \; \term{num} \\
&|& \ts{Ts} \\
&|& \ts{Sint} \\
&|^r& \nterm{kind} \ts{-!>} \nterm{kind} \\
&|& \ts{*} \ts{[} \nterm{kind}^*_{\ts{,}} \ts{]} \\
\end{gramrule}
\subsection{Miscellanenous}
\begin{gramrule}
\nterm{convar} &::=& \term{ident} \\
\nterm{label} &::=& \term{ident} \\
\end{gramrule}
\section{Context Sensitive Checks}
\paragraph{Dword ptr checks}  \mbox{}\ts{dword ptr} may be used only before a
memory operand, that is, an operand of the form $\ts{[}\ldots\ts{]}$.
\ts{dword ptr} must be used before memory operands involving labels rather than
registers; \ts{dword ptr} must be used in unary operands for the instructions
\ts{dec}, \ldots, \ts{rcl}, \ldots, \ts{call}, \ts{jmp}, \ts{pop}, and
\ts{push}; \ts{dword ptr} must be used in binary operands and the \ts{mov}
instruction where the other operands is an immediate, that is, number or
singleton.
\section{Type Checking}
\subsection{Kinds}
\subsubsection{Well Formedness}
\newcommand{\tjk}[1]{\vdash#1}
All kinds are well formed: \[\infer{\tjk{k}}{}\]
\subsubsection{Subkinding}
\newcommand{\tjsk}[2]{\vdash#1\le#2}
\[
\infer{\tjsk{\kappa}{\kappa}}{\tjk{\kappa}}
\qquad
\infer{\tjsk{\ts{T1}}{\ts{T}}}{}
\qquad
\infer{\tjsk{\ts{T2}}{\ts{T}}}{}
\qquad
\infer{\tjsk{\ts{T4}}{\ts{T}}}{}
\qquad
\infer{\tjsk{\ts{T8}}{\ts{T}}}{}
\qquad
\infer{\tjsk{\ts{Tm}\;n}{\ts{Tm}}}{}
\]
\[
\infer{\tjsk{\kappa_{11}\ts{-!>}\kappa_{12}}{\kappa_{21}\ts{-!>}\kappa_{22}}}
      {\tjsk{\kappa_{21}}{\kappa_{11}} & \tjsk{\kappa_{12}}{\kappa_{22}}}
\qquad
\infer{\tjsk{\ts{*}\ts{[}\kappa_{11}\ts{,}\ldots\ts{,}\kappa_{1n}\ts{]}}
            {\ts{*}\ts{[}\kappa_{21}\ts{,}\ldots\ts{,}\kappa_{2n}\ts{]}}}
      {\tjsk{\kappa_{1i}}{\kappa_{2i}}}
\]
It follows that kinds ordered by subkinding form a partial order and
that if two kinds have a lower bound or upper bound they have a meet
or join respectively.
\subsection{Type Constructors}
\subsubsection{Well Formedness}
\newcommand{\tjt}[3]{#1\vdash#2:#3}
\[
\infer[(\Delta(\alpha)=\kappa)]{\tjt{\Phi;\Delta}{\alpha}{\kappa}}{}
\qquad
\infer[(\Phi(\ell)=\kappa)]{\tjt{\Phi;\Delta}{\ts{`}\ell}{\kappa}}{}
\]
\[
\infer{
\tjt{\Phi;\Delta}
    {\ts{fn}\;\alpha_1\ts{:}\kappa_1\;\ldots\;\alpha_n\ts{:}\kappa_n\;\ts{.}
       \;c}
    {\kappa_1\ts{-!>}\cdots\ts{-!>}\kappa_n\ts{-!>}\kappa}
}{
\tjt{\Phi;\Delta,\alpha_1:\kappa_1,\ldots,\alpha_n:\kappa_n}{c}{\kappa}
}
\]
\[
\infer{\tjt{\Phi;\Delta}{c_1\;c_2}{\kappa}}
      {\tjt{\Phi;\Delta}{c_1}{\kappa_1\ts{-!>}\kappa} &
       \tjt{\Phi;\Delta}{c_2}{\kappa_2} &
       \tjsk{\kappa_2}{\kappa_1}}
\]
\[
\infer{\tjt{\Phi;\Delta}{\ts{[}c_1\ts{,}\ldots\ts{,}c_n\ts{]}}
                        {\ts{*}\ts{[}\kappa_1\ts{,}\ldots\ts{,}\kappa_n\ts{]}}}
      {\tjt{\Phi;\Delta}{c_i}{\kappa_i}}
\qquad
\infer[(0\le i<n)]
      {\tjt{\Phi;\Delta}{c\ts{.}i}{\kappa_i}}
      {\tjt{\Phi;\Delta}{c}
                        {\ts{*}\ts{[}\kappa_1\ts{,}\ldots\ts{,}\kappa_n\ts{]}}}
\]
\subsubsection{Subtyping}
\newcommand{\tjst}[3]{#1\vdash#2\le#3}
\end{document}
% EOF: talx86_lang_spec.tex
