\documentclass{article}

\title{An Overview of TALx86}
\author{Greg Morrisett, Karl Crary, Neal Glew, \& Dave Walker \\
	Cornell University}

\begin{document}
\maketitle
\section{Introduction}

TALx86 is a typed version of Intel's IA32 ({\em i.e.}, the 32-bit,
flag segment subset of the 80x86 architecture.)  The type system
allows programmers and compilers to build strong abstractions at the
level of assembly language.  In this respect, TALx86 is similar to
Java VM bytecodes -- we can compile high-level languages to it
and still verify that the target code respects the source-level
typing abstractions.  However, unlike Java VM, there is no need
for a trusted just-in-time (JIT) compiler.  In particular, the
only thing necessary to turn a TALx86 program into an executable
is an assembler (notably Microsoft's MASM version 6.11) and a linker
(notably Microsoft's Visual C++ linking tool).  In future
releases, we expect to provide tools for manipulating typed
{\em binaries} or DLLs, thereby avoiding the need for an assembler.

Included in our distribution are tools for parsing, printing, and
type-checking TALx86 assembly code.  We've also included two very
simple, prototype compilers that generate TALx86 code.  The first
compiler, {\tt popcorn}, compiles a type-safe subset of a C-like
programming language, and includes support for features such as arrays
and structs, separate compilation, abstract data types, {\em etc.}
The second compiler, {\tt mini-scheme}, compiles a (dynamically-typed)
lambda-calculus based language similar to Scheme.

This document attempts to explain the TALx86 type system and how
to use it as a target programming language.  We use the {\tt popcorn}
and {\tt mini-scheme} compilers to demonstrate how various language
features may be compiled to assembly language in a type-preserving
manner.


\end{document}